Spending too much time at the Galactic bar: Chaotic evolution of the Ophiuchus stream

### 2015-11-26 ms.tex
working on intro

The Ophiuchus stream \citep{bernard14, sesar15a} is a recently discovered stellar tidal stream that sits above the Galactic bulge at a Galactocentric radius and height $(R,z) \approx (1, 4.3)~{\rm kpc}$. All observational evidence suggests that the stream is a totally disrupted globular cluster: The stream stars have (1) a small positional dispersion orthogonal to the extended direction of the stream ($\approx$10 pc); (2) a small velocity dispersion $\approx$0.4 ${\rm km}~{\rm s}^{-1}$; (3) an old stellar population ($\approx$12 Gyr) estimated from isochrone fitting; and (4) no detectable over-density along the stream that could be the progenitor system \citep[][hereafter S15a]{sesar15a}.

The deprojected length of the visible part of the stream ($\approx$1.5 kpc) is short given the age of its stellar population. S15a fit an orbit to the kinematics of the stream stars in a static, axisymmetric model for the gravitational field of the inner Galaxy and ran N-body simulations of globular clusters on this orbit. S15a find that---on this orbit---the observed portion of the stream must have been formed in the last $\lesssim$400 Myr for the stream to remain as short as it is observed. This dynamical age is at odds with the old stellar population. More recently, a significant observational effort has followed-up nearly all blue horizontal branch (BHB) star candidates within XX deg of the stream to obtain radial velocities to identify more stream members \citep[][hereafter S15b]{sesar15b}. The stream has a very distinct and large line-of-sight velocity ($\approx$290 \kms) and is therefore easily detected above the background halo population. Of the \todo{XX} stars followed up, six have velocities significantly discrepant with the halo population. Of these six stars, two lie within the previously detected extent of the stream and have velocities consistent with the previously identified stream members. The remaining four stars lie close to an extrapolation of the stream, have velocities much larger than halo stars in this region ($v_{\rm los}>230$ \kms), but have a velocity dispersion $\approx$75 times the measured internal velocity dispersion of the stream stars and positional spread much larger than the orthogonal positional dispersion of the main part of the stream. These low-density, high-dispersion features were not modeled in S15a and N-body simulations on the orbit fit from the previous work do not  reproduce these features.

It is not surprising that the orbit fit in the static, axisymmetric potential model used in S15a leads to a dynamical puzzle and failed to predict the new features: it is now well-known that the Galactic bulge contains a triaxial, bar-like structure several kpc in size \citep[e.g.,][]{blitzXX, wegg13, MANY}. Given the proximity of the stream to the center of the Galaxy, the time-dependent, triaxial potential of the Galactic bar must be taken into account when modeling the orbit of the progenitor system. The presence of a bar-like perturbation to the potential will change the possible orbit of the stream progenitor and will drastically change the resonance structure of orbits in the inner galaxy \citep{athanassoula, portail15b, zotos12}, potentially introducing a significant number of chaotic orbits in this region \citep{weinberg15}.

@@

The mean orbits in the barred potentials are all strongly chaotic. We have tried computing the frequency diffusion rate for these orbits but have found that, over two consecutive integration windows, the frequency recovery fails because the frequency spectrum changes dramatically. Over timescales of $\approx$10--20 orbital periods.

% \section{Results ??: Characterizing chaos ..?}
%Using the orbit initial conditions described above (Section~\ref{sec:orbit-grids}), we integrate orbits and compute the frequency diffusion rate for the static potential and for XX barred models with varied initial bar angle and pattern speed. We find that [...]

% Full table with everything - boring
%\begin{table*}[ht]
%\footnotesize
%\begin{center}
%	\begin{tabular}{ccccccccc}
%	\toprule
%	name & $\phi_2$ [deg] & $d$ [kpc] & $\mu_l$ [mas yr$^{-1}$] & $\mu_b$ [mas yr$^{-1}$] & $v_r$ [km s$^{-1}$] & $s_{\phi_2}$ [deg] & $s_{d}$ [kpc] & $s_{v_r}$ [km s$^{-1}$]\\\midrule
%static & $-0.03\pm0.05$ & $8.35\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.1\pm0.9$ & $0.20\pm0.04$ & $0.31\pm0.10$ & $2.9\pm0.8$\\
%bar1 & $-0.03\pm0.05$ & $8.40\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.0\pm0.9$ & $0.21\pm0.05$ & $0.31\pm0.14$ & $3.2\pm0.8$\\
%bar2 & $-0.03\pm0.05$ & $8.40\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $288.9\pm1.1$ & $0.20\pm0.04$ & $0.33\pm0.09$ & $3.3\pm0.9$\\
%bar3 & $-0.04\pm0.06$ & $8.40\pm0.06$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.0\pm1.0$ & $0.21\pm0.05$ & $0.35\pm0.09$ & $3.2\pm0.8$\\
%bar4 & $-0.03\pm0.06$ & $8.41\pm0.06$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.1\pm1.0$ & $0.21\pm0.05$ & $0.36\pm0.10$ & $3.4\pm0.9$\\
%bar5 & $-0.03\pm0.05$ & $8.41\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.1\pm0.9$ & $0.21\pm0.04$ & $0.35\pm0.09$ & $3.3\pm0.8$\\
%bar6 & $-0.03\pm0.05$ & $8.40\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.0\pm1.0$ & $0.21\pm0.05$ & $0.35\pm0.10$ & $3.2\pm0.8$\\
%bar7 & $-0.04\pm0.05$ & $8.41\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.0\pm1.0$ & $0.21\pm0.05$ & $0.33\pm0.12$ & $3.4\pm0.9$\\
%bar8 & $-0.03\pm0.05$ & $8.42\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $289.1\pm1.0$ & $0.20\pm0.04$ & $0.34\pm0.09$ & $3.4\pm0.9$\\
%bar9 & $-0.02\pm0.05$ & $8.41\pm0.05$ & $-7.2\pm0.1$ & $0.9\pm0.1$ & $288.9\pm1.0$ & $0.21\pm0.04$ & $0.35\pm0.09$ & $3.3\pm0.8$\\
%	\bottomrule
%	\end{tabular}
%	\caption{Estimated mean and standard deviation of the marginal posterior distributions over each parameter in our orbit fit likelihood. All values are nearly identical because the potentials are very similar over the short length of the stream. This table is mainly included as a sanity check---we have made samples from the full posterior distribution available with this article and provide code to transform to and from stream coordinates (see http://adrian.pw/ophiuchus/coordinates for more information).\label{tbl:param-means} }
%\end{center}
%\end{table*}

@@

@@

>>>>

The Ophiuchus stream \citep{bernard14, sesar15a} is a recently discovered stellar tidal stream that sits above the Galactic bulge at a Galactocentric radius and height $(R,z) \approx (1, 4.3)~{\rm kpc}$. All observational evidence suggests that the stream is a totally disrupted globular cluster: The stream stars have (1) a small positional dispersion orthogonal to the extended direction of the stream ($\approx$10 pc); (2) no detectable over-density along the stream that could be the progenitor system; (3) a small velocity dispersion $\approx$0.4 ${\rm km}~{\rm s}^{-1}$; and (4) an old stellar population ($\approx$12 Gyr) estimated from isochrone fitting \citep[][hereafter S15a]{sesar15a}.

There are a number of peculiarities about the observed kinematics of probable member stars. For example, the de-projected length of the visible part of the stream is short given the age of its stellar population ($\approx$1.5 kpc). S15a fit an orbit to the kinematics of the stream stars in a static, axisymmetric model for the gravitational field of the inner Galaxy and ran N-body simulations of globular clusters on this orbit. S15a find that---on this orbit---the portion of the stream visible as an over-density in main-sequence stars must have been formed in the last $\lesssim$400 Myr for the stream to remain as short as it is observed. This dynamical age is at odds with the old stellar population. Another puzzle is the existence of blue horizontal branch (BHB) stars close to the stream (within XX deg) with similar radial velocities, but with a large dispersion in both sky position and velocity \citep[][hereafter S15b]{sesar15b}. The stream has a very distinct and large line-of-sight velocity ($\approx$290 \kms) and is therefore easily detected above the background halo population. Four stars have been detected with line-of-sight velocities $>230~\kms$ that lie close to an extrapolation of the stream on the sky, but have a velocity dispersion $\approx$75 times larger than the measured internal velocity dispersion of the stream stars. These stars hint at the existence of associated low-density, high-dispersion features that were not modeled in S15a and are not predicted by the $N$-body simulations from this prior work.

The orbit fit and $N$-body simulations in S15a used a static, axisymmetric potential to represent the Milky Way potential, but it is now well-known that the Galactic bulge contains a triaxial, bar-like structure several kpc in size \citep[e.g.,][]{blitzXX, wegg13, MANY}. Given the proximity of the stream to the center of the Galaxy, the time-dependent, triaxial potential of the Galactic bar must be taken into account when modeling the orbit of the Ophiuchus stream. The presence of a bar-like perturbation to the potential will change the orbit of the stream progenitor and the orbit structure in the inner galaxy \citep{athanassoula, portail15b, zotos12}. Bar-like features can also introduce a significant number of chaotic orbits in their vicinity\citep{weinberg15}.

@@

The mean orbits in the barred potentials are all strongly chaotic. We have tried computing the frequency diffusion rate for these orbits but have found that, over two consecutive integration windows, the frequency recovery fails because the frequency spectrum changes dramatically over timescales of $\approx$10--20 orbital periods.

@@

The model streams presented here do not reproduce all observed features of the Ophiuchus stream. Instead, these results illustrate that chaotic evolution of tidal debris can plausibly explain the peculiar features of the stream. If the cluster progenitor was on a regular orbit, it would have to have disrupted entirely within the last $\approx$300 Myr (Figure~\ref{TODO}) in order to explain the shortness and density profile. In addition, the four most recently identified BHB stars with similar distances and line-of-sight velocities would have to be (highly unlikely) chance alignments of halo stars. If instead the progenitor were on a chaotic orbit (because of the influence of the Galactic bar), (1) stars stripped early will have ``fanned'' out and would thus be harder to observe and (2) the nearby, high-velocity-dispersion BHB stars can be naturally explained by this chaotic stream-fanning. We consider the second scenario to be more plausible: our understanding of the formation and evolution of the Ophiuchus stream is that the progenitor object has been orbiting and steadily losing stars over the last several Gigayears, but only the stars stripped from the most recent few pericentric passages remain coherent enough to be detected as a stream-like over-density in the PS1 data.

@@

[Gaia, Don't do distance cut that Brani did, chemistry]

[explored in future work: what are implications of the existence of this structure for total population]
